Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-05 main.Rmd
finished intro:

Phytopathogens are a major threat to global crop production.  The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of ~6 billion USD to potato production, annually [@haas2009genome]. The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp]. The incidences of crop disease are increasing, global climate change and agricultural practice are expanding the geographical range of pathogens and upping the stakes in the evolutionary arms race.  Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage. Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are **notoriously difficult to characterise from sequence data**. In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc]. Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb]. To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a critical need to identify effectors computationally from genome and protein sequence data.

Deep learning models can uncover cryptic patterns in data. We believe such models can be trained to identify effectors directly from sequence as others have shown it is possible to classify other eukaryote and bacterial nuclear proteins [@Savojardo:2017eq]. In doing so these models encode information about the features it is using to classify that can be extracted to reveal novel sequence features that current techniques cannot. Deep learning models are distinct from other machine learning processes in that pre-selection of important features is far less critical and the models can learn these features unsupervised from training data [@Jurtz:2017cr]. This property removes the need to know which properties of a data set must be examined before data collection begins.

Convolutional neural networks (CNNs) ...

Recent work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.  The work shows clearly that there are unknown patterns in the sequence data that a deep learning method could use to classify.

Here we use combinations of Convolutional Neural Networks (CNNs) that do not rely so heavily on _a priori_ feature selection - only a well annotated data set,  and that can extract novel features of the sequence in an unbiased, more unsupervised manner.

Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as effector or not. The larger and more accurate the list the more sensitivity a model can obtain. It is critical that training examples are experimentally verified effectors. Much of the effector annotation in public genomics databases is from computational predictions of genomics and therefore unverified hypothetical effectors. A good source of experimentally verified data is in the Molecular Plant Microbe Interactions (MPMI) literature.  The widest ranging manual curation of MPMI papers is being performed as part of the PHI-Base [@Urban:2017gp] database strategy.

@@

We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.\@ref(fig:confusionfinalmodels) The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors. The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors. The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct. Overall the models are accurate and show little to no bias with the exception of the GRU-Embedding type. In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.

@@

```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
# knitr::include_graphics("figures/cf_other_tools.png")

@@

```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
# knitr::include_graphics("figures/convolution_maps_fixed.pdf")

@@

```{r fungiefftable}

>>>>

Phytopathogens are a major threat to global crop production.  The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of ~6 billion USD to potato production, annually [@haas2009genome]. The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp]. The incidences of crop disease are increasing, global climate change and agricultural practice are expanding the geographical range of pathogens and upping the stakes in the evolutionary arms race.  Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage. Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are notoriously difficult to characterise from sequence data. In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc]. Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb]. To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a need to identify effectors computationally from genome and protein sequence data.

Machine learning (ML) algorithms are a general group of techniques most often used for classification of data into groups. Supervised ML require a set of training examples and associated data with which to learn. Defining the best data to use and collect, called feature selection is an important and difficult prerequisite.  ML approaches have been applied with success to biological sequence analysis, particularly in transcription factor binding site prediction, for the classification of eukaryote and bacterial nuclear proteins [@Savojardo:2017eq] and in the plant pathogen domain work by Sperschneider et al [@Sperschneider:2018gc] developed two ensemble-based machine learning models that could identify effectors and predict localisation with > 70% accuracy.

Deep learning models are distinct from other machine learning processes in that pre-selection of important features is far less critical and the models can learn these features unsupervised from training data [@Jurtz:2017cr]. This property removes the need to know which properties of a data set must be examined before data collection begins. The Deep learning models can therefore classify on properties not necessarily known to the operator and could be used to uncover cryptic patterns in data. Convolutional neural networks (CNNs) are a type of neural network that have found wide application in numerous machine vision problems, including image object classification and facial identification [@krizhevsky2012, @lawrence1997], in time-series data analysis [@pyrkov2018] and natural language processing [@collobert2008]. In the biomedical domain they have been used in drug discovery [@wallach2015] and gene network prediction [@maclean2019]. CNNs encode information about the features used to classify that can be extracted and interpreted. In a sequence classification problems this means they have the potential to reveal novel sequence features that other bioinformatics approaches have not and could be of particular utility when analysing sets of effectors.

Deep learning approaches require positive and negative examples from which to learn - here a list of sequences annotated as an effector or not. The larger and more accurate the list the more sensitivity a model can obtain. It is critical that training examples are experimentally verified effectors. Much of the effector annotation in public genomics databases is from computational predictions of genomics and is therefore of experimentally unverified hypothetical effectors. A good source of experimentally verified data is in the Molecular Plant Microbe Interactions (MPMI) literature and The widest ranging manual curation of MPMI papers is being performed as part of the PHI-Base [@phibase] database strategy, PHI-base is an expertly curated database of genes proven experimentally to affect the outcome of pathogen host interactions and is therefore an excellent source of reliable effector sequences.

Here we use combinations of CNNs that do not rely on _a priori_ feature selection to classify experimentally verified effectors and non-effectors in three taxa of plant pathogen: bacterial, fungal and oomycete. We show that these have very strong predictive power, can outperform existing effector prediction methods in accuracy and a better balance of sensitivity and specificity.  We also analyse the activations of the models in response to effectors and non-effectors to gain insights into the sequence features that are allowing classification. We have produced an R package that will allow other scientists to easily classify their own sequences of interest.

@@

We examined the tendency of the models to call false positives or false negatives preferentially by creating confusion matrices of the classifications relative to the ground truth on the hold-out test data.\@ref(fig:confusionfinalmodels) The bacterial sequence trained models in general showed high accuracy and only one or two miscalls with no error bias except for the GRU-Embedding model which called five from 38 effectors as non effectors. The fungal sequence trained models were less accurate overall and showed a small amount more bias, again in the GRU-Embedding model, which was biased towards calling effectors as non-effectors and the CNN-LSTM model which was slightly biased in the opposite direction, calling non-effectors as effectors. The oomycete models were again quite balanced but the GRU-Embedding model showed a quite conservative tendancy calling 12 out of 17 effectors as non-effectors whilst getting all 17 non-effectors correct. Overall the models are accurate and show little to no bias toward false positive or false negatives, with the exception of the GRU-Embedding type. In oomycete sequences in particular and in this class of model across the different sequence types showed itself to tend to call real effectors as not of that class.

@@

```{r vsotherscfmat, fig.width=9, fig.height=8, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences.", message = FALSE, echo=FALSE}
# knitr::include_graphics("figures/cf_other_tools.png")

@@

```{r activations, fig.height=6, fig.width=10, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
# knitr::include_graphics("figures/convolution_maps_fixed.pdf")

@@

```{r fungiefftable, message = FALSE, echo=FALSE}
